An assessment of the biodiversity - ecosystem function relationship in southern African woodlands

### 2019-11-13 analysis/output/include/clust_summ.tex
Made tikz path diagrams

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Thu, Oct 24, 2019 - 13:20:25
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{5pt}} ccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
clust5 & c\_ind & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
\hline \\[-1.8ex] 
C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 16+12.6 & 281+373.8 & 34.7+37.1 \\ 
C2 & Julbernardia.paniculata, Isoberlinia.angolensis, Albizia.antunesiana & 22+13.7 & 274+246.9 & 46.7+35.07 \\ 
C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 8+6.5 & 272+535.3 & 36+27.93 \\ 
C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 15+9.2 & 448+391.9 & 46.6+45.41 \\ 
C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 9+9.4 & 336+434.3 & 43.8+34.45 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

>>>>

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Fri, Nov 08, 2019 - 11:40:45
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{5pt}} ccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
clust5 & c\_ind & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
\hline \\[-1.8ex] 
C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7+4.7 & 276+345.2 & 35.3+37.03 \\ 
C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8+4.3 & 274+245.9 & 46.8+35.05 \\ 
C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 4+2.3 & 274+536.5 & 36.3+27.86 \\ 
C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 10+5.9 & 447+386.3 & 47+45.41 \\ 
C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 4+2.9 & 322+390.1 & 43.8+34.45 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-11-13 manuscript/chapter_1.tex
Made tikz path diagrams

\usepackage{dcolumn}
\usepackage{multirow}

\title{A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{John L. Godlee}
\date{}

\input{include/n_plots}

@@

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests of Costa Rica. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as much of a role in the relationship between biodiversity and ecosystem function. Instead, stress tolerance and productivity of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{}.

In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity. This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}. Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}. Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016}. The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred.

The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}. Climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in dry woodlands, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

@@

@@

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{}). We investigated the relationship between aboveground woody biomass and tree biodiversity. We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature. We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type clusters \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality. Initially, we made four hypotheses:

@@

\begin{enumerate}
	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species and structural tree diversity.}
	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1983}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.

@@

After the initial plot data cleaning described above, a Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each plot to identify compositional outliers, which were considered not representative southern African woodlands. Plots within the upper 95th percentile of neighbour distance in the PCoA were excluded from further analyses (\autoref{pcoa_outlier}). This procedure removed 66 plots.

\begin{figure}[H]
\centering
	\includegraphics[width=0.5\textwidth]{pcoa_outlier}
	\caption{Principle Co-ordinate Analysis of tree species diversity of the \nplotspcoa{} plots chosen from the larger pool of 5395 plots. The PCoA is constructed from a tree species abundance matrix. Plots which are within the upper 95th percentile of plot neighbour euclidean distance in the PCoA are marked in red, and were excluded from further analysis.}
	\label{pcoa_outlier}
\end{figure}

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. AGB estimation for stems <5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}. Additionally, for some species, species identification is difficult for saplings.

All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem, species, DBH and tree height to the top of the highest branch material were recorded. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause potentially large measurement uncertainty in the dataset. Averaged over such a large number of stems however, this should not meaningfully alter the results of our analysis. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from \citet{Ryanunbublished}.

@@

Climatic data was collected from the ECMWF ERA5 dataset \citep{}. Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of these variables across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil. These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Rarefied species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region. Larger plots will tend to encompass more individuals, and therefore probabilistically more species, according to the species-area relationship \citep{}.

@@

Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the rarefied and extrapolated number of tree species in the plot.

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}. A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration. Independently of total precipitation, precipitation seasonality, measured as the covariance of daily precipitation, determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{con_mod}
	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity, and AGB. Hypothesised paths of causal inference are depicted as arrows from predictor to response. Latent variables are represented as shaded boxes with their constituent observed variables within. The names of each latent variable are in ovals. The pale orange dashed box defines a nested model which includes only tree diversity variables.}
	\label{con_mod}
\end{figure}

Observed variables were standardized to Z-scores for the SEMs. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}.

@@

Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability, and soil fertility, on AGB both directly and via species diversity.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation.}
	\label{corr_mat}
\end{figure}

At the regional level, tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB, while environmental observed variables had weak correlations with AGB. Only MAP and soil fertility observed variables (CEC, organic C \%, sand \%) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero. Stronger correlations were seen between certain environmental covariates and measures of tree species diversity, namely, MAP, MAT, soil organic C \%, and soil sand \% (\autoref{corr_mat}).

\subsection{Structural diversity and species diversity models}

In an SEM describing the effect of tree species diversity on AGB via the mediating effect of stand structural diversity (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}). Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}), but this effect was not as strong as the effect of species diversity itself. Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).

Model fit was also good when the model was refitted separately using data from each of the 5 vegetation clusters. It should be noted that model fit was not as good as when using regional level data, which may have arisen due to the small sample sizes available for some vegetation clusters, most notably C4, which only had 46 plots. While the strength of the path coefficients varied between vegetation clusters, the sign of the path coefficients remained positive and significant in all cases except for the direct effect of diversity on biomass in cluster 3 (\autoref{struc_model_slopes_all}).

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{struc_mod}
	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
	\label{struc_mod}
\end{figure}

@@

@@

\subsubsection{Mediation by stem density}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line charts showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
	\label{sem_struc_stems_ha}
\end{figure}

\subsection{Models with environmental covariates}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_mod}
	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
	\label{full_mod}
\end{figure}

@@

\input{include/full_model_fit_clust_stats}

\subsection{Structural diversity}

The latent variable of tree species diversity had a positive effect on AGB across the region. A stronger positive effect was seen between structural diversity and aboveground biomass, with species diversity acting \textit{through} structural diversity to effect aboveground biomass.

\subsection{Interaction effects of moisture availability and soil}

@@

@@

\bibliography{/Users/johngodlee/google_drive/bib/lib}

>>>>

\usepackage{dcolumn}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}

\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{John L. Godlee}
\date{}

\input{include/n_plots}
\input{include/n_outliers}
\input{include/hull_cover}
\input{include/path_coef_struc}
\input{include/path_coef_full}

@@

The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems. \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage. In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}. In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as much of a role in the relationship between biodiversity and ecosystem function. Instead, stress tolerance and productivity of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}. Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{}.

\todo{Re-assess this paragraph -}In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity. This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}. Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}. Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}. Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016}. The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle. Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}. These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}. Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}. A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred. \todo{Say how the previous studies might have been wrong because of their limited scope, or how in order to really answer the question you need to sample long environmental gradients.}

The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}. In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas. \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum. However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.

@@

\todo{Stand structural complexity}

@@

In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{}). We investigated the relationship between aboveground woody biomass and tree biodiversity. We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature. We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass. We incorporated vegetation type clusters based on tree species composition \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality. Initially, we made four hypotheses:

@@

\begin{enumerate}
	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.}
	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
\end{enumerate}

@@

The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1983}). The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique. The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}). The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.

@@

After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition. We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers. We set the threshold value to remove the top 5\% of plots with the highest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.

@@

We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{}. Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses. Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems forming a very small proportion of the total AGB in a plot \citep{}.

All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements. A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual. For each stem, species, DBH and tree height to the top of the highest branch material were recorded. Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause varying measurement uncertainty by data source in the dataset. When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m. The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from \citet{Ryanunbublished}.

@@

Climatic data was collected from the ECMWF ERA5 dataset \citep{}. Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot. Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data. Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}. We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil. These data are a modelled product compiled from various remote sensed and directly measured data sources.

@@

\subsection{Data analysis}
Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}. This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value. Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers. Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region. Larger plots will tend to encompass more individuals, and therefore probabilistically more species \citep{}.

@@

Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.

We calculated tree stand structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.

@@

Structural Equation Models (SEM) investigated the determinants of AGB. All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R}. SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}. A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression. Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset. For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration. Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.

@@

\begin{figure}[H]
\centering
	\input{include/path_diagram_concept}
	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity, and AGB. Hypothesised paths of causal inference are depicted as arrows from predictor to response. Latent variables are represented as shaded boxes with their constituent observed variables within. The names of each latent variable are in ovals. The pale orange dashed box defines a nested model which includes only tree diversity variables.}
	\label{con_mod}
\end{figure}

Observed variables were standardized to Z-scores for the SEMs. Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one. Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}. Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}. To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1. For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed. Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.

@@

Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.6\textwidth]{corr_mat}
	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation. Variables in bold were later reversed for modelling to maintain positive correlations between observed variables within each latent variable. Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero. Colours of variable names group variables into latent variables: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity.}
	\label{corr_mat}
\end{figure}

A simple correlation plot showed that all tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB. Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51). Environmental variables had weaker correlations with AGB than diversity variables. Only MAP ($\rho$ = 0.19), MAP seasonality ($\rho$ = -0.18), soil Organic C \% ($\rho$ = 0.19) and soil Sand \% ($\rho$ = -0.18) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero. (\autoref{corr_mat}).

As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB. MAT and the seasonality of temperature variation did not correlate with AGB. As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.

MAP had positive correlations with tree species diversity, tree height diversity and a positive but not significant correlation with tree stem density. MAT had non significant correlations with tree species and structural diversity variables.

Tree species diversity variables had clear positive correlations with stem density (Species richness: $\rho$ = 0.45, Shannon equitability: $\rho$ = 0.43).

\subsection{Structural and species diversity models}

In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}). Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}). Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).

When the model was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but generally similar relationships between diversity and AGB with the same sign and significant overlap between the 95\% confidence intervals of path coefficients. It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).

\begin{figure}[H]
\centering
	\input{include/path_diagram_struc}
	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
	\label{struc_mod}
\end{figure}

%Meaasurement errors of exogenous variables omitted for clarity

@@

\subsection{Variation among vegetation clusters in structural and species diversity effects}

@@

\subsubsection{Moderation of Diversity-AGB relationship by stem density}

@@

\begin{figure}[H]
\centering
	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
	\label{sem_struc_stems_ha}
\end{figure}

\subsection{Mediation of environmental covariates via diversity}

@@

\begin{figure}[H]
\centering
	\input{include/path_diagram_full}
	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
	\label{full_mod}
\end{figure}

@@

% \input{include/full_model_fit_clust_stats} - Excluded, low fit

Vegetation cluster level models could not be reliably fitted for these more complex models with environmental covariates, due to sample size and convergence issus.

\subsection{Moderation effects of moisture availability and soil}

@@

subsection{Large trees}

We analysed how the strength of the relationship between tree species richness and AGB varied for different size classes of tree. \autoref{bchave_quantile_slope_sp_rich} shows a saturating increase in the importance of species richness to determine AGB as the proportional size of trees increases.

@@

\subsection{Effect of tree species diversity on AGB}

\subsection{Structural diversity as a mechanism for the BEFR}

\subsection{Moderation effects of moisture availability and soil fertility}

\subsection{Vegetation type specific responses}

\subsection{Stem density effects}

\todo{Why is there a peak in the effect of species richness on AGB at a certain level of stem density?}

%\bibliography{/Users/johngodlee/google_drive/bib/lib}
